quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-10-19 paper-quollr.tex
Added implementation workflow

The implementation of \texttt{quollr} is designed to be efficient, and easy to extend. The package is organized into a series of logical components that reflect the main stages of the workflow: data preprocessing, model fitting, low-density bin removal, prediction, visualization, and interactive exploration. This package structure makes the code easier to maintain and allows new features to be added without disrupting the existing functionality.

@@

Internally, \texttt{quollr} is organized into five primary modules that correspond to key steps in the analysis pipeline:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Data preparation --- functions such as \texttt{gen\_scaled\_data()} standardize the input, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
\item
  Hexagonal binning --- the \texttt{hex\_bin\_data()} module constructs a two-dimensional hexagonal grid and aggregates local summaries of the high-dimensional data for each bin.
\item
  Model fitting --- the \texttt{fit\_highd\_model()} function computes high-dimensional centroids and pairwise bin connections to build a structured manifold representation.
\item
  Error computation --- functions like \texttt{compute\_errors()} and \texttt{aggregate\_bin\_error()} evaluate local and global discrepancies between the model and data, enabling diagnostic visualization.
\item
  Visualization and linkage --- plotting and tour functions (\texttt{show\_hex\_model()}, \texttt{show\_link\_plots()}, and \texttt{show\_error\_link\_plots()}) integrate outputs into linked static and dynamic displays using \texttt{ggplot2}, \texttt{langevitour}, and \texttt{crosstalk}.
\end{enumerate}

@@

The internal data objects in \texttt{quollr} adhere to the \textbf{tidy data} principle, ensuring that every variable forms a column, every observation forms a row, and every type of observational unit forms a table.
This structure supports compatibility with the \texttt{tidyverse} and other R visualization frameworks.

The main object classes include:

\begin{itemize}
\tightlist
\item
  \textbf{\texttt{highd\_data}} -- a tibble containing the original high-dimensional observations with a unique identifier (\texttt{ID}) and variable columns prefixed with \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, \ldots).
\item
  \textbf{\texttt{nldr\_data}} -- a tibble containing two-dimensional embeddings, labeled as \texttt{emb1} and \texttt{emb2}, matched to the same \texttt{ID}s.
\item
  \textbf{\texttt{highd\_vis\_model}} -- the primary model object returned by \texttt{fit\_highd\_model()}, containing two key components: the high-dimensional summaries for each bin and their corresponding two-dimensional positions.
\item
  \textbf{\texttt{hex\_bin\_obj}} -- the internal representation of the hexagonal grid, including bin coordinates (\texttt{a1}, \texttt{a2}), centroids, and normalized weights (\texttt{w\_h}).
\end{itemize}

These objects provide a coherent foundation for data transfer across \texttt{quollr}'s modules---from binning to model fitting and visualization---supporting reproducibility and transparency in analyses.

@@

\begin{verbatim}
bin_configs <- calc_bins_y(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 25, 
  q = 0.1)

@@

\begin{verbatim}
> $b2
> [1] 36
> 
> $a1
> [1] 0.04943533
> 
> $a2
> [1] 0.04281225
\end{verbatim}

@@

\begin{verbatim}
hb_obj <- hex_binning(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 25, 
  q = 0.1)
\end{verbatim}

@@

\begin{verbatim}
all_centroids_df <- gen_centroids(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 25, 
  q = 0.1
  )

all_centroids_df
\end{verbatim}

\begin{verbatim}
> # A tibble: 900 x 3
>        h      c_x    c_y
>    <int>    <dbl>  <dbl>
>  1     1 -0.1     -0.127
>  2     2 -0.0506  -0.127
>  3     3 -0.00113 -0.127
>  4     4  0.0483  -0.127
>  5     5  0.0977  -0.127
>  6     6  0.147   -0.127
>  7     7  0.197   -0.127
>  8     8  0.246   -0.127
>  9     9  0.295   -0.127
> 10    10  0.345   -0.127
> # i 890 more rows
\end{verbatim}

@@

\begin{verbatim}
>   h           x           y
> 1 1 -0.10000000 -0.09858838
> 2 1 -0.12471766 -0.11285913
> 3 1 -0.12471766 -0.14140063
> 4 1 -0.10000000 -0.15567138
> 5 1 -0.07528234 -0.14140063
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>    emb1  emb2    ID     h
>   <dbl> <dbl> <int> <int>
> 1 0.300 0.985     1   659
> 2 0.721 0.688     2   492
> 3 0.722 0.542     3   418
> 4 0.196 1.01      4   657
> 5 0.240 1.06      5   708
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 3
>       h   n_h   w_h
>   <int> <int> <dbl>
> 1    82     2 0.002
> 2    83     1 0.001
> 3   106     1 0.001
> 4   107     5 0.005
> 5   108     7 0.007
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 2
>       h pts_list 
>   <int> <list>   
> 1    82 <int [2]>
> 2    83 <int [1]>
> 3   106 <int [1]>
> 4   107 <int [5]>
> 5   108 <int [7]>
\end{verbatim}

@@

\begin{verbatim}
>   h          c_x        c_y n_h w_h
> 1 1 -0.100000000 -0.1271299   0   0
> 2 2 -0.050564674 -0.1271299   0   0
> 3 3 -0.001129349 -0.1271299   0   0
> 4 4  0.048305977 -0.1271299   0   0
> 5 5  0.097741302 -0.1271299   0   0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>    from    to   x_from  y_from     x_to    y_to from_count to_count
>   <int> <int>    <dbl>   <dbl>    <dbl>   <dbl>      <dbl>    <dbl>
> 1     1     2 -0.1     -0.127  -0.0506  -0.127           0        0
> 2    26    27 -0.0753  -0.0843 -0.0258  -0.0843          0        0
> 3     3    27 -0.00113 -0.127  -0.0258  -0.0843          0        0
> 4    26    52 -0.0753  -0.0843 -0.0506  -0.0415          0        0
> 5    52    53 -0.0506  -0.0415 -0.00113 -0.0415          0        0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 10
>    from    to   x_from  y_from     x_to    y_to from_count to_count
>   <int> <int>    <dbl>   <dbl>    <dbl>   <dbl>      <dbl>    <dbl>
> 1     1     2 -0.1     -0.127  -0.0506  -0.127           0        0
> 2    26    27 -0.0753  -0.0843 -0.0258  -0.0843          0        0
> 3     3    27 -0.00113 -0.127  -0.0258  -0.0843          0        0
> 4    26    52 -0.0753  -0.0843 -0.0506  -0.0415          0        0
> 5    52    53 -0.0506  -0.0415 -0.00113 -0.0415          0        0
> # i 2 more variables: from_reindexed <int>, to_reindexed <int>
\end{verbatim}

@@

\begin{verbatim}
low_density_hex <- find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 25, 
  md_thresh = 0.05
)
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>       h    x1    x2    x3       x4        x5       x6        x7
>   <int> <dbl> <dbl> <dbl>    <dbl>     <dbl>    <dbl>     <dbl>
> 1    82 0.989  1.78  1.13  0.00470  0.0142   -0.0547  -0.00428 
> 2    83 0.960  1.93  1.28  0.00278 -0.0180   -0.0678  -0.00260 
> 3   106 0.997  1.37  1.07  0.00154  0.00687   0.0289   0.00760 
> 4   107 0.975  1.46  1.20 -0.00497 -0.000595  0.0139   0.000537
> 5   108 0.931  1.64  1.36 -0.00706  0.00487  -0.00705  0.000383
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>   pred_emb_1 pred_emb_2    ID pred_h
>        <dbl>      <dbl> <int>  <int>
> 1      0.295      0.986     1    659
> 2      0.716      0.686     2    492
> 3      0.740      0.558     3    418
> 4      0.197      0.986     4    657
> 5      0.246      1.07      5    708
\end{verbatim}

>>>>

The implementation of \texttt{quollr} is designed to be efficient, and easy to extend. The package is organized into a series of logical components that reflect the main stages of the workflow: data preprocessing, model fitting, low-density bin removal, prediction, visualization, and interactive exploration (Figure \ref{fig:workflow}). This package structure makes the code easier to maintain and allows new features to be added without changing the existing functionality.

\begin{figure}[H]
\includegraphics[width=1\linewidth]{figures/quollr_workflow} \caption{Workflow for }\label{fig:workflow}
\end{figure}

@@

The package is organized into seven core modules corresponding to stages of the analysis workflow: preprocessing, $2\text{-}D$ model construction, lifting into $p\text{-}D$, prediction, error computation, visualization, and interactivity. Each module performs a distinct task and communicates through data objects.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Data preprocessing --- The function \texttt{gen\_scaled\_data()} standardizes the embedding data, manage variable naming, and ensure consistent identifiers across high-dimensional and embedded datasets.
\item
  Construct $2\text{-}D$ model --- \texttt{hex\_binning()}, \texttt{merge\_hexbin\_centroids()}, \texttt{tri\_bin\_centroids()}, \texttt{gen\_edges()}, \texttt{update\_trimesh\_index()}
\item
  Lift the model into $p\text{-}D${} --- the \texttt{avg\_highd\_data()}
\item
  Prediction - \texttt{predict\_emb()}
\item
  Error computation --- \texttt{glance()}
\item
  Visualization --- \texttt{geom\_hexgrid()}, \texttt{geom\_trimesh()}, \texttt{show\_langevitour()}
\item
  Interactivity - \texttt{show\_link\_plots()}, and \texttt{show\_error\_link\_plots()}
\end{enumerate}

@@

The internal data objects follow the tidy data principle: each variable is stored in a column, each observation in a row, and each type of information in its own table. This structure makes the package easy to use with the \texttt{tidyverse} and other R visualization tools.

\subsubsection{Input objects}\label{input-objects}

\begin{itemize}
\item
  \texttt{highd\_data} -- a tibble containing the original high-dimensional observations with a unique identifier (\texttt{ID}) and variable columns prefixed with \texttt{"x"} (e.g., \texttt{x1}, \texttt{x2}, \ldots).
\item
  \texttt{nldr\_data} -- a tibble containing two-dimensional embeddings, labeled as \texttt{emb1} and \texttt{emb2}, matched to the same \texttt{ID}s.
\end{itemize}

\subsubsection{Generated objects}\label{generated-objects}

\begin{itemize}
\item
  \texttt{scaled\_nldr\_obj} --- the output of \texttt{gen\_scaled\_data()}, which rescales the embedding to the range \([0, 1] \times [0, y_{2,\text{max}}]\), where \(y_{2,\text{max}} = r_2 / r_1\) is the ratio of the embedding ranges. It includes the scaled coordinates (\texttt{scaled\_nldr}) and the original limits (\texttt{lim1}, \texttt{lim2}).
\item
  \texttt{hex\_bin\_obj} -- the object created by \texttt{hex\_binning()}, which defines the structure of the two-dimensional hexagonal grid used in modeling. It includes the grid spacing (\texttt{a1}, \texttt{a2}), the number of bins along each axis, the centroids of all hexagons, polygon coordinates for plotting, and the mapping of each data point to its assigned hexagon.
\item
  \texttt{highd\_vis\_model} --- the main model object returned by \texttt{fit\_highd\_model()}. It stores all components of the fitted visualization model, including the scaled NLDR data (\texttt{nldr\_scaled\_obj}), the hexagonal bin structure (\texttt{hb\_obj}), the averaged $p\text{-}D$ summaries for each bin (\texttt{model\_highd}), the corresponding $2\text{-}D$ bin centroids (\texttt{model\_2d}), and the triangulated mesh connecting neighboring bins (\texttt{trimesh\_data}).
\end{itemize}

@@

\begin{verbatim}
bin_configs <- calc_bins_y(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 21, 
  q = 0.1)

@@

\begin{verbatim}
> $b2
> [1] 30
> 
> $a1
> [1] 0.05966332
> 
> $a2
> [1] 0.05166995
\end{verbatim}

@@

\begin{verbatim}
hb_obj <- hex_binning(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 21, 
  q = 0.1)
\end{verbatim}

@@

\begin{verbatim}
all_centroids_df <- gen_centroids(
  nldr_scaled_obj = scurve_umap_obj, 
  b1 = 21, 
  q = 0.1
  )

head(all_centroids_df, 5)
\end{verbatim}

\begin{verbatim}
> # A tibble: 5 x 3
>       h     c_x    c_y
>   <int>   <dbl>  <dbl>
> 1     1 -0.1    -0.127
> 2     2 -0.0403 -0.127
> 3     3  0.0193 -0.127
> 4     4  0.0790 -0.127
> 5     5  0.139  -0.127
\end{verbatim}

@@

\begin{verbatim}
>   h           x           y
> 1 1 -0.10000000 -0.09268324
> 2 1 -0.12983166 -0.10990656
> 3 1 -0.12983166 -0.14435320
> 4 1 -0.10000000 -0.16157652
> 5 1 -0.07016834 -0.14435320
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>    emb1  emb2    ID     h
>   <dbl> <dbl> <int> <int>
> 1 0.300 0.985     1   448
> 2 0.721 0.688     2   351
> 3 0.722 0.542     3   287
> 4 0.196 1.01      4   468
> 5 0.240 1.06      5   489
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 3
>       h   n_h   w_h
>   <int> <int> <dbl>
> 1    49     1 0.001
> 2    68     1 0.001
> 3    69     8 0.008
> 4    70     5 0.005
> 5    89     3 0.003
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 2
>       h pts_list 
>   <int> <list>   
> 1    49 <int [1]>
> 2    68 <int [1]>
> 3    69 <int [8]>
> 4    70 <int [5]>
> 5    89 <int [3]>
\end{verbatim}

@@

\begin{verbatim}
>   h         c_x        c_y n_h w_h
> 1 1 -0.10000000 -0.1271299   0   0
> 2 2 -0.04033668 -0.1271299   0   0
> 3 3  0.01932665 -0.1271299   0   0
> 4 4  0.07898997 -0.1271299   0   0
> 5 5  0.13865330 -0.1271299   0   0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>    from    to  x_from  y_from    x_to    y_to from_count to_count
>   <int> <int>   <dbl>   <dbl>   <dbl>   <dbl>      <dbl>    <dbl>
> 1     1     2 -0.1    -0.127  -0.0403 -0.127           0        0
> 2    22    23 -0.0702 -0.0755 -0.0105 -0.0755          0        0
> 3    22    44 -0.0702 -0.0755 -0.0403 -0.0238          0        0
> 4     3    23  0.0193 -0.127  -0.0105 -0.0755          0        0
> 5    44    45 -0.0403 -0.0238  0.0193 -0.0238          0        0
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 10
>    from    to  x_from  y_from    x_to    y_to from_count to_count from_reindexed
>   <int> <int>   <dbl>   <dbl>   <dbl>   <dbl>      <dbl>    <dbl>          <int>
> 1     1     2 -0.1    -0.127  -0.0403 -0.127           0        0              1
> 2    22    23 -0.0702 -0.0755 -0.0105 -0.0755          0        0             22
> 3    22    44 -0.0702 -0.0755 -0.0403 -0.0238          0        0             22
> 4     3    23  0.0193 -0.127  -0.0105 -0.0755          0        0              3
> 5    44    45 -0.0403 -0.0238  0.0193 -0.0238          0        0             44
> # i 1 more variable: to_reindexed <int>
\end{verbatim}

@@

\begin{verbatim}
low_density_hex <- find_low_dens_hex(
  model_2d = df_bin_centroids, 
  b1 = 21, 
  md_thresh = 0.05
)
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 8
>       h    x1    x2    x3        x4       x5        x6        x7
>   <int> <dbl> <dbl> <dbl>     <dbl>    <dbl>     <dbl>     <dbl>
> 1    49 0.999  1.93  1.05  0.00276   0.0133  -0.0530   -0.00270 
> 2    68 0.997  1.37  1.07  0.00154   0.00687  0.0289    0.00760 
> 3    69 0.960  1.58  1.27 -0.00709   0.00759 -0.000368 -0.000292
> 4    70 0.935  1.81  1.35 -0.000338 -0.00713 -0.0278   -0.00474 
> 5    89 0.985  1.13  1.14 -0.00597  -0.00830  0.0196   -0.00140
\end{verbatim}

@@

\begin{verbatim}
> # A tibble: 5 x 4
>   pred_emb_1 pred_emb_2    ID pred_h
>        <dbl>      <dbl> <int>  <int>
> 1      0.318      1.01      1    470
> 2      0.735      0.700     2    351
> 3      0.705      0.545     3    287
> 4      0.198      1.01      4    468
> 5      0.228      1.06      5    489
\end{verbatim}
